Relation of severe COVID-19 to household composition, recent hospital exposure, and advice to shield in clinically vulnerable and low-risk individuals in Scotland: REACT-SCOT  case-control studyDeterminants of rates of serious hypoglycaemic events in type 1 diabetes

### 2021-01-09 shielding.Rmd
updated shielding paper, fixed trainmodel.R

---
title: "Risk of COVID-19 in clinically extremely vulnerable individuals advised to shield during the epidemic: the REACT-SCOT case-control study" 
author:
  - name: Paul M McKeigue
    affiliation: "HPS,Usher"
  - name: David Caldwell 
    affiliation: HPS
  - name: Ciara Gribben
    affiliation: HPS
  - name: Jen Bishop
    affiliation: HPS
  - name: Matthew Armstrong
    affiliation: HPS
  - name: Joke Delvaux
    affiliation: HPS
  - name: Sharon Hutchinson 
    affiliation: GCU
  - name: David McAllister 
    affiliation: Glasgow  
  - name: Chris Robertson 
    affiliation: Strathclyde
  - name: Nazir Lone
    affiliation: Usher
  - name: Janet Murray
    affiliation: HPS 
  - name: Jim McMenamin  
    affiliation: HPS
  - name: Alison Smith-Palmer 
    affiliation: HPS
  - name: David Goldberg
    affiliation: HPS
  - name: Helen M Colhoun
    affiliation: "HPS,IGMM,NHSFife"
  - name: Public Health Scotland COVID  -19 Health Protection Study Group
    affiliation: HPS

address: 
- code: IGMM
  address: Institute of Genetics and Molecular Medicine, College of Medicine and Veterinary Medicine, University of Edinburgh, Western General Hospital Campus, Crewe Road, Edinburgh EH4 2XUC, Scotland. HC - AXA Chair in Medical Informatics and Epidemiology
- code: Usher
  address: Usher Institute, College of Medicine and Veterinary Medicine, University of Edinburgh, Teviot Place, Edinburgh EH8 9AG, Scotland. PM - Professor of Genetic Epidemiology and Statistical Genetics. NL - Clinical Senior Lecturer in Critical Care  
- code: HPS
  address: Health Protection Scotland (Public Health Scotland), Meridian Court, 5 Cadogan Street, Glasgow G2 6QE
- code: Glasgow
  address: Institute of Health and Wellbeing, University of Glasgow, 1 Lilybank Gardens, Glasgow G12 8RZ. DM - Senior Clinical Lecturer in Public Health 
- code: Strathclyde
  address: Department of Mathematics and Statistics, University of Strathclyde, 16 Richmond Street, Glasgow G1 1XQ. CR - Professor of Public Health Epidemiology
- code: GCU
  address: School of Health and Life Sciences, Glasgow Caledonian University. SH - Professor of Epidemiology and Population Health 
- code: ISD
  address: Information Services Division, NHS Scotland, UK

header-includes:
  \usepackage{newunicodechar}
  \DeclareUnicodeCharacter{039B}{\ensuremath{\mit\Lambda}}
  \let\origquote\quote
  \def\quote{\origquote\itshape}
  \usepackage{graphicx}
  \usepackage{geometry}
  \usepackage{longtable}
  \usepackage{booktabs}
  \usepackage{float}
  \usepackage{array}
  \usepackage{threeparttable}
  \usepackage{longtable}
  \usepackage{lscape}
  \usepackage{pdfpages}
  \usepackage[nolists, nomarkers]{endfloat}

always_allow_html: true
urlcolor: blue
output:
      bookdown::pdf_book:  
        base_format: rticles::elsevier_article
        keep_tex: true
        fig_caption: yes

geometry: "left=2cm,right=2cm,top=2cm,bottom=2cm"

bibliography: ./covid.bib
csl: ./biomed-central.csl
linenumbers: false
linkcolor: cyan

@@

Prepared for the Shielding Policy Evaluation team and the Clinical Leads for Shielding Group (CLAGS)

Correspondence: Professor Helen Colhoun 07791652665/ helen.colhoun@igmm.ed.ac.uk

```{r functions, echo=FALSE}
library(knitr)
library(kableExtra)
library(bookdown)

options(knitr.kable.NA = '')
options(knitr.table.format = "latex") # adds tab: prefix to labels
knitr::opts_chunk$set(echo = FALSE, message=FALSE, warning=FALSE)
source("helperfunctions.R")

@@

@@

# Background  
The SARS-CoV-2 pandemic reached Scotland in early 2020 with the first positive laboratory test being recorded on March 1st. Extreme social distancing measures were soon imposed, as in many other countries, with subsequent relaxation now that the effective reproduction number of cases is below 1. The overall policy is one of suppression and if monitoring indicates a second wave of COVID-19, more stringent measures are likely to be be re-imposed.

Early in the epidemic it was realised that certain people would likely be extremely vulnerable to severe COVID-19 disease including those receiving certain immunosuppressant drugs or with other conditions.  Therefore Public Health Scotland  instituted a process of identifying such persons through their primary care physician and through searching of administrative databases  (see appendix for full protocol). Those considered eligible then received a letter detailing the need for self isolation and the support they would receive to achieve this. These letters were sent out over several weeks between xx-and xx.

An initial survey of ~12000 shielding eligible persons indicates that in general some degree of compliance was high but often partial and it varied across different socio-demographic groups.

Here we take the original listing of ~ 180,228 persons deemed eligible for shielding at the outset of the epidemic. We examine the cumulative incidence of any evidence of COVID-19 and any serious COVID-19. using a case control design of all serious COVID-19 cases and age and sex matched controls in Scotland we examine the odds of disease associated with being shielding eligible. We explore whether there is any evidence that shielding reduced risks in these persons.

Using data from the period in which shielding was then stopped ( 1 August 2020) and the next instruction issued to take extra  precautions we examine the number of cases that have occurred.

# Methods
The design of the REACT-SCOT case-control study has been described in detail elsewhere [@mckeigue_rapid_2020-1].

## Ascertainment of eligibility for shielding

To identify  all those nationally deemed eligible for shielding we took the list generated by PHS at the outset of the epidemic (see https://hpspubsrepo.blob.core.windows.net/hps-website/nss/3008/documents/1_covid-19-search-criteria-highest-risk-patients.pdf  for details) .

In brief the following groups were identified

* Solid organ transplant recipients

* People with specific cancers

* People with severe respiratory conditions: cystic fibrosis, severe asthma, severe chronic obstructive airway disease, on home oxygen, severe bronchiectasis, pulmonary hypertension)

* People with rare diseases

* People on immunosuppression therapies

* People who are pregnant with significant heart disease, congenital or acquired.

* People on renal dialysis and who have had a splenectomy were also added ** to which group?

* A 7th group for primary and secondary care clinicians to identify patients who should be advised to shield.

 As noted in the search criteria document cited above patients were identified via searches of hospital discharge and prescribing records, as well as other sources including regional cancer chemotherapy databases, NHS Blood and Transplant and individual specialist clinician lists.

The list contained the CHI number which enabled linkage to other datasets. it also contained# the reason for eligibility and the date on which the shielding letter was sent to the person concerned.

## COVID-19 case ascertainment and definition
For the total Scottish population "Evidence of  any detected COVID-19" was obtained as having had a positive  nucleic acid test  for SARS-CoV-2 in  the national virology laboratory database,  or having had a hospital discharge code for COVID-19 in those with no positive test, or a death with COVID-19 mention anywhere on the death certificate even where testing had not been  positive.  The databases used were  the Electronic Communication of Surveillance in Scotland (ECOSS) database  that captures all virology testing in all NHS laboratories nationally,  the National Records of Scotland (NRS) death registrations data, the RAPID
database which is a daily update of hospitalisation nationwide and Scottish Morbidity Records-01 which captures hospital discharges  and discharge diagnostic codes nationally.

Databases were queried from March 1st 2020  (date of the first diagnosed case of COVID-19 in Scotland) up to the `r format(max(cc.all$SPECIMENDATE), "%d %b %Y")` for test results and `r format(max(cc.all$DATE_OF_REGISTRATION, na.rm=TRUE), "%d %b %Y")` for death registrations.

All these databases use the  Community Health Index (CHI) identifier. The CHI number is a unique identifier for everyone in the population, and is used in all care systems in Scotland. The CHI database is the master database for the CHI number and  contains several fields including age, sex, residential postcode, and residential care home status.

Among all those with any evidence of detected COVID-19 we then defined severe disease as that requiring critical care or  fatal.   All admissions to intensive (ICU), high dependency (HDU) or combined ICU/HDU care (hereafter critical care unit-CCU) were obtained by linkage using the CHI number  to the Scottish Intensive Care Society and Audit Group (SICSAG) database. Fatal COVID-19 was defined  based on their being a mention of  COVID-19 anywhere  on the death certificate regardless of prior testing or not, OR death within 28 days of testing positive  (or of the admission date for those SMR01 ascertained cases) regardless of whether COVID-19 was mentioned on the death certificate or not. Note that these definitions of COVID-19 deaths are the official ones used by the  National Records of Scotland and Public Health Scotland, respectively.

## Ascertainment of controls
For every case i.e. person with "any evidence of COVID-19" "as defined above, we selected 10 controls matched for sex, one year age band and attending the same GP practice  and alive on the same date of the index case , test, two weeks prior to index case admission or two weeks prior to death for cases ascertained through testing, SMR01 admission data, or NRS death data respectively.  The sampling frame for controls was the Community Health Care Index.

## Demographic data
For all those on the shielding list   and for all COVID-19 cases and controls i)socioeconomic status was evaluated using the postcode-based indicator the Scottish Index of Multiple Deprivation (SIMD) using the residential postcode in the CHI database [@fraser_scottish_2020]. Postcodes were allocated to which quintile of the socioeconomic distribution of Scottish residential datazones they belonged.  (ii) Ethnic group
was captured by linkage to prior SMR01 and SMR00 and SMR04  records   in which the
patients self assigns ethnicity according to the groups that are used in the
national census. iii) Residential care home status for those with diabetes was available via the
care home flag for postal address in the CHI database.

## Other morbidity and drug prescribing
For all those on the shielding list   and for all COVID-19 cases and controls, so as to ascertain other comorbidities the  general hospital admissions data in the  Scottish Morbidity Record 01 (SMR01),  and  all discharge codes over the past five years and Anatomical Therapeutic Chemical codes [@who_anatomical_2016] from Prescribing Information Systems (PIS) data for the past three years were extracted and used to define co-morbid conditions and prior drug exposures .

We first extracted history of a specific list of conditions and drug classes that have been included as moderate risk conditions for COVID-19 by public health agencies hereafter termed "listed conditions" [@nhs_whos_2020]. These co-morbid conditions were assigned based on either a diagnostic code in any hospital discharge record, or a dispensed prescription in which that diagnostic code is the only indication (see Supplementary Materials for details).

## Statistical methods
\begin{itemize}

\item First we tabulated the shielding list by age-group  sex and reason  for shielding.

\item We then tabulated the \% ( and CI for the \%) of those on the shielding list with "Any evidence of COVID-19" as described above by age and sex   and reason for shielding as well as overall.

\item We also tabulated the \% of those on the shielding list with "Severe i.e. CCU treated or fatal COVID-19 " as described above, by age and sex   and reason for shielding as well as overall.

\item To compare risks in shielded persons before and after shielding was introduced For each shielded person we  marked up their longitudinal record  from Feb 7th with date of shielding letter, date of stop shielding letter (Aug 1st)  and any date of death with the record being censored at data of death. We then computed the number of cases of any evidence of COVID-19  per total  person days in pre shielded time, then shielded time , then post shielded time.  We repeated this for severe cases.

\item Among the shielding to understand whether there are other determinants of becoming a case, we compared the  \% with each of the  other listed conditions,  the \% with more than 1 listed condition, the number of prescribed ATC drug classes, the \% in  residential care home,  the  SIMD  distribution and ethnicity  distribution in those who became cases and those who did not.

\item To provide a comparator for the background population,  we used  the 2019 age band and sex - mid year estimates for the general population and the shielding status in cases to derive  the \% ( and 95\%CI) with any evidence of COVID-19 and severe COVID-19 in the same age and sex groups in the general population  (DAVID this is available from STUART). We then used logistic regression of the one year age band counts of cases and non cases in the shielded and general population to compute and OR associated with shielding.

\item The odds of "exposure "to eligibility for shielding in cases  versus  controls was also  quantified as the conditional odds ratio  from a conditional logistic regression.  We did this because it was easy to extend this regression was then extended to include SIMD, ethnicity, residential care home status,  the "listed "COVID-19 risk conditions.

\item Since the calculation in 4. above can be confounded by overall background trends in the Re to further evaluate if there is any evidence that shielding reduced risk, The odds of "exposure "to eligibility for shielding in cases  versus  controls  BY WEEK of epidemic was also quantified as the conditional odds ratio  from a conditional logistic regression.

\end{itemize}

Analyses were conducted using the R statistical package .

# Results

Table \ref{tab:shieldgroups} shows the frequency of shielding categories. The 6 women pregnant with heart disese were allocated to the "clinician identified" category.

Table \ref{tab:letterdates} shows a breakdown by date of sending letter

Table \ref{tab:afterletter} shows cases among shielded individuals classified by whether estimated date of infection was before or after the shielding letter.

Table \ref{tab:regression.allcases} shows the estimated rate ratios for any diagnosis of COVID-19

Table \ref{tab:regression.severecases} shows the estimated rate ratios for severe COVID-19.

# Discussion

## Summary of findings

## Comparison with previous literature

## Limitations and advantages 
Some key advantages of our study are the population coverage, the range of
outcomes assessed, the extensive prior risk factor and clinical history
data.

## Policy Implications

# Declarations
## Information governance

>>>>

---
title: "Relation of severe COVID-19 to household composition, recent hospital exposure, and advice to shield in clinically vulnerable and low-risk individuals in Scotland: REACT-SCOT  case-control studyDeterminants of rates of serious hypoglycaemic events in type 1 diabetes"
author:
  - name: Paul M McKeigue
    affiliation: "HPS,Usher"
  - name: David McAllister 
    affiliation: Glasgow 
  - name: David Caldwell 
    affiliation: HPS
  - name: Ciara Gribben
    affiliation: HPS
  - name: Jen Bishop
    affiliation: HPS
  - name: Matthew Armstrong
    affiliation: HPS
  - name: Joke Delvaux
    affiliation: HPS
  - name: Sam Colville
    affiliation: HPShttps://t.co/ro5FIUsaWr?amp=1
  - name: Sharon Hutchinson 
    affiliation: GCU
  - name: Chris Robertson 
    affiliation: Strathclyde
  - name: Nazir Lone
    affiliation: Usher
  - name: Jim McMenamin 
    affiliation: HPS
  - name: David Goldberg
    affiliation: HPS
  - name: Helen M Colhoun
    email: helen.colhoun@igmm.ed.ac.uk (07791 652665)
    affiliation: "HPS,IGMM,NHSFife"

address: 
- code: Usher
  address: Usher Institute, College of Medicine and Veterinary Medicine, University of Edinburgh, Teviot Place, Edinburgh EH8 9AG, Scotland. PM - Professor of Genetic Epidemiology and Statistical Genetics. NL - Clinical Senior Lecturer in Critical Care  
- code: IGMM
  address: Institute of Genetics and Molecular Medicine, College of Medicine and Veterinary Medicine, University of Edinburgh, Western General Hospital Campus, Crewe Road, Edinburgh EH4 2XUC, Scotland. HC - AXA Chair in Medical Informatics and Epidemiology
- code: HPS
  address: Public Health Scotland, Meridian Court, 5 Cadogan Street, Glasgow G2 6QE
- code: Glasgow
  address: Institute of Health and Wellbeing, University of Glasgow, 1 Lilybank Gardens, Glasgow G12 8RZ. DM - Senior Clinical Lecturer in Public Health 
- code: Strathclyde
  address: Department of Mathematics and Statistics, University of Strathclyde, 16 Richmond Street, Glasgow G1 1XQ. CR - Professor of Public Health Epidemiology
- code: GCU
  address: School of Health and Life Sciences, Glasgow Caledonian University. SH - Professor of Epidemiology and Population Health 
- code: ISD
  address: Information Services Division, NHS Scotland, UK

header-includes:
  \usepackage{newunicodechar}
  \let\origquote\quote
  \def\quote{\origquote\itshape}
  \usepackage{graphicx}
  \usepackage{geometry}
  \usepackage{longtable}
  \usepackage{booktabs}
  \usepackage{float}
  \floatplacement{figure}{H}
  \usepackage{array}
  \usepackage{threeparttable}
  \usepackage{longtable}
  \usepackage{lscape}
  \usepackage{pdfpages}
output:
  rticles::nejm_article:
    includes:
      in_header: ./preamble.tex
    latex_engine: lualatex
    keep_tex: true
    fig_caption: yes
#    number_sections: false
    md_extensions: -autolink_bare_uris
geometry: "left=2cm,right=2cm,top=2cm,bottom=2cm"
bibliography: ./covid.bib
csl: ./biomed-central.csl
always_allow_html: true
urlcolor: blue
linenumbers: false
linkcolor: cyan

@@

On behalf of Public Health Scotland COVID-19 Health Protection Study Group

Prepared for the Shielding Policy Evaluation team and the Clinical Leads for Shielding Group

```{r functions, echo=FALSE}
library(knitr)
library(kableExtra)

options(knitr.kable.NA = '')
options(knitr.table.format = "latex") # adds tab: prefix to labels
knitr::opts_chunk$set(echo = FALSE, message=FALSE, warning=FALSE)
knitr::opts_chunk$set(fig.width=6, fig.pos = "H", out.extra = "")
source("helperfunctions.R")

@@

dateformat <- function(x) { # format date and strip leading zero
     gsub("^0", "", format(x, "%d %B %Y"))
}

format.estcipv <- function(x.ci, x.pval) {
    x <- gsub(" \\(", " \\(95\\% CI ", as.character(x.ci))
    x <- gsub(", ", " to ", x)
    x <- gsub("\\)", paste0(", _p_=\\", x.pval, "\\)"), x)
    x <- gsub("times", "\\\\times", x)
    return(x)
}

peakdate <- casedates.allgr[which.max(avdaily), date]
maxdeaths <- casedates.allgr[which.max(avdaily), avdaily]
deaths.quartermax.date <- casedates.allgr[date > peakdate & avdaily < 0.25 * maxdeaths,
                                 date][1]

peakdate.elig <- casedates.byelig[listedgr3=="Eligible for shielding"][which.max(avdaily), date]
maxdeaths.elig <- casedates.byelig[listedgr3=="Eligible for shielding"][which.max(avdaily), avdaily]
deaths.elig.quartermax.date <- casedates.byelig[listedgr3=="Eligible for shielding" &
                                         date > peakdate.elig &
                                         avdaily < 0.25 * maxdeaths.elig,
                                         date][1]

@@

 \newpage

# Abstract
* _Objectives_ - to investigate the relation of severe COVID-19 to factors influencing exposure to SARS-CoV-2 and the implications of this for protecting vulnerable individuals through shielding and targeted vaccination.

* _Design_ - (1) cohort study of all individuals identified as eligible for shielding; (2) matched case-control study (REACT-SCOT).

* _Setting_ - population of Scotland from 1 March 2020 to 15 October 2020

* _Participants_ - The cohort study comprised `r nrow(shielded.full)` individuals identified as eligible for shielding by Public Health Scotland. The case-control study included  all `r nrow(cc.all[CASE==1])` diagnosed cases of COVID-19 and `r nrow(cc.all[CASE==0])` controls matched for age, sex and primary care practice.

* _Interventions_ - Letters advising eligible individuals to shield were sent by Public Health Scotland from 3 April 2020 onwards.

* _Main outcome measure_ - severe COVID-19 defined as cases that entered critical care or were  fatal.

* _Results_ - `r nrow(shielded.full[!is.na(casegr2) & casegr2=="Severe"])` individuals eligible for shielding developed severe COVID-19. Although the incidence of severe disease fell from early April 2020 onwards in the Scottish population, the rate ratio associated with eligibility for shielding increased.  With those without risk conditions as reference category, the univariate rate ratio for severe disease was
`r format.estcipv(table.listedgr3.severecases.nocare["Moderate risk condition", 3], table.listedgr3.severecases.nocare["Moderate risk condition", 4])` 
in those with moderate risk conditions and
`r format.estcipv(table.listedgr3.severecases.nocare["Eligible for shielding", 3], table.listedgr3.severecases.nocare["Eligible for shielding", 4])` 
in those eligible for shielding.  The highest rate ratio for severe disease was `r format.estcipv(table.shielded.severecases.nocare["Solid organ transplant", 3], table.shielded.severecases.nocare["Solid organ transplant", 4])` in solid organ transplant recipients.  In all risk groups the strongest risk factors (in terms of deviance explained) for severe COVID-19 were recent hospital exposure with a rate ratio of `r format.estcipv(table.shielded.severecases.nocare["Recent hospital visit/stay", 3], table.shieldedonly.severecases.nocare["Recent hospital visit/stay", 4])` and number of adults in the household, with a rate ratio of  `r format.estcipv(table.listedgr3.severecases.nocare["adultsgt1", 3], table.listedgr3.severecases.nocare["adultsgt1", 4])` for two or more adults compared with one.

* _Conclusions_ -  Protection of vulnerable individuals requires control of transmission in hospital settings and prevention of transmission from other adults in the household.  For solid organ transplant recipients, in whom vaccination is likely to be ineffective, vaccination of others in the household should not be ruled out as a means of reducing transmission.

# Background  
The SARS-CoV-2 pandemic reached Scotland in early 2020 with the first positive laboratory test recorded on 1 March 2020. Population-wide interventions included advice that symptomatic individuals should self-isolate, banning of mass gatherings, closure of schools and finally a population-wide lockdown on 23 March 2020.  Although no systematic studies of risk factors were available at that time, public health agencies issued lists of "moderate risk" conditions and "diseases and conditions considered to be very high risk".  Those with "very high risk" conditions were designated as clinically extremely vulnerable and eligible for shielding.

The practicality of shielding vulnerable individuals, in the context of a population-wide lockdown where contacts have been reduced to a low level, depends on the extent to which their exposure to likely sources of infection either directly or through other household members can be reduced. It is now possble in Scotland to study this through record linkage to household composition, recent hospital exposure, and with relevant occupations including health care [@shah_risk_2020] and teaching [@mcallister_report_2020].

The objectives of this study were to investigate the relation of severe COVID-19 to factors influencing exposure to SARS-CoV-2 and to examine the implications of this for protecting vulnerable individuals through shielding and targeted vaccination.

# Methods
## Ascertainment of eligibility for shielding 
On 21 March 2020 a [list](https://www.england.nhs.uk/coronavirus/wp-content/uploads/sites/52/2020/03/200321_GP-letter_Annex-2_FINAL_2.pdf) of "diseases and conditions considered to be very high risk" was issued by NHS England.  Six categories were initially designated as eligible for shielding [@health_protection_scotland_search_2020]:

1. Solid organ transplant recipients

2. People with cancer of the blood or bone marrow at any stage of treatment, or people with cancer receiving treatments that affect the immune system

3. People with severe respiratory conditions including cystic fibrosis, severe asthma and severe chronic obstructive airway disease, on home oxygen, severe bronchiectasis, pulmonary hypertension)

4. People with rare diseases that increase the risk of infections such as severe combined immunodeficiency, homozygous sickle cell disease.

5. People on immunosuppression therapies sufficient to increase risk of infection

6. Women who were pregnant with heart disease

An additional category comprised patients identified by clinicians as requiring shielding advice.  On 11 May 2020 the list of eligible conditions was updated to include those on renal dialysis and those who had had a splenectomy.

Public Health Scotland (PHS) began identifying patients in this category through health-care information systems including hospital discharge records, prescription encashments, regional cancer chemotherapy databases, blood and transplant registries, and direct requests to clinicians in relevant specialties [@health_protection_scotland_search_2020]. Those identified as eligible were sent a letter advising them to self-isolate and offering support.  The first batch of letters was sent on `r dateformat(batches$Date.Sent)[1]`.

The cohort of `r nrow(shielded.full)` individuals comprised the list generated by Public Health Scotland at the outset of the epidemic, updated with those persons newly identified as eligible up to the end of July. Letters were sent on Fridays, beginning on `r dateformat(batches$Date.Sent[1])` an throughout the first wave until `r  dateformat(max(shielded.full$Date.Sent, na.rm=TRUE))` though advice to shield was rescinded on 1 August 2020.

Those identified as eligible were advised to "work from home if possible" and that employers should "support you to do this2 or "make sure suitable arrangements are in place so you can go to work."  Help was offered with deliveries of food and medicine, but no support was offered for others resident in the household eitheir to isolate with the vulnerable individual or to move out temporarily.

## COVID-19 case ascertainment and definition
Case ascertainment has been described in detail elsewhere [@mckeigue_rapid_2020-1].  Case ascertainment was based on querying the following national-level databases: Electronic Communication of Surveillance in Scotland (ECOSS  that captures virology testing in all NHS laboratories, National Records of Scotland (NRS) death registrations,  RAPID which is a daily update of hospitalisations, and Scottish Morbidity Record 01 (SMR01) which records general /acute hospital discharges including day cases. All these databases use the Community Health Index (CHI) number as identifier.  The CHI database includes age, sex, postcode and care home status, and can be queried to extract numbers of adults and children in the household.

Cases of COVID-19 were ascertained as those with a positive  nucleic acid test for SARS-CoV-2 in ECOSS, a hospital discharge diagnosis of COVID-19 in SMR01, or a death registration with mention of COVID-19 anywhere on the death certificate.  The presentation date was assigned as the date of the first positive test for those ascertained through testing, as seven days before the admission date for those without a positive test result ascertained through hospital discharge records, and as fourteen days before the date of death for those without a positive test result ascertained through death certificates. Databases were queried from 1 March 2020  (date of the first diagnosed case of COVID-19 in Scotland) up to `r format(max(cc.all$SPECIMENDATE), "%d %b %Y")` for test results and `r format(max(cc.all$DATE_OF_REGISTRATION, na.rm=TRUE), "%d %b %Y")` for death registrations.

Severe COVID-19 was defined by entry to critical care within 21 days of presentation date, or fatal outcome. Entry to critical care units -- intensive care, high dependency or combined units -- was obtained by linkage to the Scottish Intensive Care Society and Audit Group (SICSAG) database. Fatal outcome was defined as death with  COVID-19 coded as underlying cause, death from any cause within 28 days of testing positive, or death within 28 days of presentation date for cases ascertained through discharge records.  This definition of severe cases was chosen to ensure that case ascertainment would not be affected by triage of those assessed as unlikely to benefit from critical care, or by the scale-up of SARS-CoV-2 testing.

In cases with fatal outcome, median time from presentation to death was `r with(cc.all[fatalcase==1], median(as.integer(Date.Death -SPECIMENDATE)))` days. The proportion of cases with fatal outcome was calculated over sliding seven-day time windows.  We denote this fraction as the fatal case proportion, rather than the case fatality ratio which is classically defined without linkage between diagnoses and deaths, as the ratio of deaths from the disease to diagnosed cases over a specified time window.

As those who were notified as having died were removed from the list of those eligible for shielding before letters were sent out, and no records of this filtering were available, we examined the fatal case proportion by presentation date.  The rationale for this was that as filtering to remove those who had died would misallocate fatal cases from those eligible for shielding to those ineligible, we could estimate the date up to which this misallocation occurred by examining the fatal case proportion by presentation date.

## Ascertainment of controls
As linkage to demographic and morbidity data was not available for the full shielding cohort. these associations were examined in the REACT-SCOT case-control study.The design of this study has been described in detail elsewhere [@mckeigue_rapid_2020-1].   For each case of COVID-19, up to 10 controls matched for sex, one-year age band and primary care practice were selected from the CHI database, and assigned the same presentation date as the case. With this incidence density sampling design, it is possible and correct for an individual to appear more than once as a control and subsequently as a case.

## Linkage of cases and controls to demographic and morbidity data
Socioeconomic status was assigned as the Scottish Index of Multiple Deprivation (SIMD) score which is based on linkage of postcodes to Census data [@fraser_scottish_2020].  Quintile 5 is the least deprived.  Care home residence was assigned using the field in the CHI database. As this does not capture all care homes, those aged over 70 in households with 10 or more adults were also coded as care home residents.  Cases and controls were linked to hospital discharge codes over the last five years in SMR01, to BNF codes of dispensed prescriptions in the 240 days before presentation date in the Prescribing Information System (PIS) and to the national register of diabetes. As described previously  [@mckeigue_rapid_2020-1], we used these linked datasets to assign a list of "moderate risk conditions" for COVID-19 designated by public health agencies [@nhs_whos_2020].  These are referred to as "listed conditions".  Three risk groups were defined: no risk condition, moderate risk condition but ineligible for shielding, and eligible for shielding.

To examine possible associations with exposure to health-care facilities, we defined the variable "recent hospital visit/stay" for cases and controls to include those individuals who had been hospital in-patients or attended hospital out-patient clinics at any time in the 14 days before presentation date, excluding admissions or outpatient attendances less than five days before presentation date. <!--- check whether outpatient attendances include online consultations --->  This exclusion was intended to exclude admissions caused by COVID-19 but to include anyone exposed to health care facilities in the time interval during which the infection is likely to have been acquired.  This is broader than the definition used in Scotland and England for "probable hospital-acquired infection" which is restricted to cases who had been hospital inpatients for at least seven days before they first tested positive. In contrast to this restrictive definition of probable hospital-acquired infection which identifies individuals who are unlikely to have been infected outside hospital, this inclusive definition of "recent hospital visit/stay" allows us to calculate the fraction of cases attributable to this exposure.

## Statistical methods
The case-control and time series analyses focus on severe COVID-19 as this case definition does not depend upon the extent of testing for SARS-CoV-2, which varied markedly over the study period and is also likely to differ between those eligible for shielding and those ineligible.

For the full shielding cohort, we tabulated each category of eligibility by age group, outcome and date of sending advice letter. 2-day or 3-day time windows of numbers of severe cases were calculated for the study period. To examine the effect of advice to shield, the case-control dataset was split into two intervals at a time point fourteen days after the first batch of shielding advice letters were sent on 3 April 2020:  presentation dates up to 17 April 2020 were coded as pre-advice, and presentation dates after that as post-advice.  Care home residents were excluded from these analyses as shielding advice intended for individuals in private households would not have been relevant to care home residents.

Rate ratios were estimated in the case-control study by conditional logistic regression. For time series analyses, these rate ratios were calculated over 20-day time windows.

# Results
## Cumulative risk of COVID-19 up to presentation date of `r format(max(cc.all$SPECIMENDATE, na.rm=TRUE), "%d %b %Y")` 
Supplementary Figure \ref{fig:plotcasefatality} compares the time series of the fatal case proportion in those eligible and ineligible for shielding, by seven-day window of presentation date. The fatal case proportion in those eligible for shielding reached a peak in the first week of March and fell after this, suggesting that from mid-March onwards any misallocation of fatal cases to those ineligible for shielding was minimal.

## Shielding eligibility categories and clinical outcome
Table \ref{tab:shieldgroupsbyage} shows the frequency of risk categories in those eligible for shielding. The largest risk category was those with severe respiratory disease.   Solid organ transplant recipients and those on immunosuppressants were younger than those in the other categories.

<!--- Supplement Table \ref{tab:shieldgroupsbyage} put this into supplement or  better still omit  altogether its not really that useful eyond being brodly sensible and not showing any emajor problem with classfcaiton but it stll doesnt say anythng about sensitvity or specifcty of the shelding assignation  shows a validation check of whether the SMR01 and PIS  histories of those n the shielded category among all cases and controls is coherent with their assigned shielding category - data frequency of designated risk conditions (ascertained from diabetes registry, hospital discharge records and dispensed prescriptions) in each shielding eligibility category.  This shows that only 840 out of 2200 individuals on immunosuppressants were identified from prescriptions in primary care, as expected because many immunosuppressant drugs are prescribed only in hospital.  However 7816 out of 8062 individuals with severe respiratory disease were identified from hospital discharge records and dispensed prescriptions.
--->

Table \ref{tab:shieldcasegroup} shows the number in each eligibility category by case status classified as not diagnosed, not severe or severe COVID-19 .  Overall among `r nrow(shielded.full)` persons eligible  for shielding `r table.shield.casegr["All shielding categories", 2]` were diagnosed with COVID-19 but did not enter critical care or die within 28 days, `r table.shield.casegr["All shielding categories", 3]` entered critical care for COVID-19 but survived, and `r table.shield.casegr["All shielding categories", 4]` had a fatal outcome. No cases were diagnosed among those identified as pregnant with heart disease. Women in this group were allocated to the "Additional conditions" category in subsequent analyses. Those on immunosuppressants had the lowest risks.

<!--I think it would be useful to show among the shielded the distribution of any evidence and of severe cases (In columns)  by  age and sex simd residential care home status  and do a logistic regresson among them  ... then I can write somehtng like this ...
Among the shielded risk  of COVID-19 varied strongly by age  and ???  adjusted by age it varied by shielding category . Adjusted for age and shielding category  and  was / was not higher in men than women, in those in residential care homes than  not   and did /di not show an association  with socioeconomic status.

** We can't show a breakdown of demographic factors on the shielding cohort as no linkage has been done 
--->

Figure \ref{fig:plotcasesbydate} compares the time series of severe cases in the three risk group as a ribbon plot.  The daily number of presentations of severe cases in the total population reached a peak on `r dateformat(peakdate)` and fell to one quarter of the peak daily number by `r dateformat(deaths.quartermax.date)`.  After the first wave of the epidemic had subsided at the end of May 2020, those eligible for shielding accounted for most of the severe cases still occurring until the second wave begain in late September 2020.

### Association of severe COVID-19 with shielding eligibility status and risk factors related to exposure
`r cc.severe[CASE==1 & shield.any==TRUE, .N, by=care.home][2, 2]` of the `r cc.severe[CASE==1 & shield.any==TRUE, .N]` severe cases and `r cc.severe[CASE==0 & shield.any==TRUE, .N, by=care.home][2, 2]` of the `r cc.severe[CASE==0 & shield.any==TRUE, .N]` matched controls were resident in care homes.  In subsequent analyses, care home residents have been excluded as policies relevant to shielding residents of care homes are different from those relevant to shielding individuals living independently.

 With those without risk conditions as reference category, the univariate rate ratio for severe disease was
`r format.estcipv(table.listedgr3.severecases.nocare["Moderate risk condition", 3], table.listedgr3.severecases.nocare["Moderate risk condition", 4])` 
in those with moderate risk conditions and
`r format.estcipv(table.listedgr3.severecases.nocare["Eligible for shielding", 3], table.listedgr3.severecases.nocare["Eligible for shielding", 4])` 
in those eligible for shielding.

Table \ref{tab:regressionseverecasesnocare} shows the rate ratios for severe COVID-19 associated with shielding eligibility categories and other relevant covariates, with care home residents excluded. Solid organ transplant recipients were the group at highest risk, with a univariate rate ratio of  `r format.estcipv(table.shielded.severecases.nocare["Solid organ transplant", 3], table.shielded.severecases.nocare["Solid organ transplant", 4])` for severe COVID-19.

The strongest risk factor (in terms of deviance explained) for severe COVID-19 in the general population was recent hospital exposure: rate ratio `r format.estcipv(table.shielded.severecases.nocare["Recent hospital visit/stay", 3], table.shielded.severecases.nocare["Recent hospital visit/stay", 4])`.

The risk of severe COVID-19 rose steeply with the number of adults in the household, but was inversely associated with the number of school-age children in the household in a  multivariable model.  The other demographic factor associated with increased risk was socioeconomic deprivation: the rate ratio in the least deprived quintile compared with the most deprived quintile was `r format.estcipv(table.shielded.severecases.nocare["5 - least deprived", 3], table.shielded.severecases.nocare["5 - least deprived", 4])`..
